Contrastive Optimization -- Preliminaries

### 2024-08-22 paper/main.tex
offloaded logging for memory optimizations; fixed leaky memory, improved writeup notation
updated plots, evals, added model weights, moved code to SGD for evaluation

\title{Interpretable Risk Minimization -- Preliminaries}

@@

Where $s_c$ denotes the logit for a single class. $\tilde{\mathcal{\bm{A}}}_c^{\text{HiResCAM}}$ generates a class-specific CAM. For the CAM architecture, \cite{draelos2020use} (section 3.6.2) shows that:
\begin{gather}
	\sum^{D_1,D_1}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} \\
	s_c = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} + b_c = \sum^{D_1,D_1}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c
\end{gather}
Therefore, we can recreate the entire input-dependent part of the score using HiResCAMs. Doing this for each class $c$, we obtain the bias-removed logit vector $\in \mathbb{R}^{c}$.

@@

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c + b_c - s_{c'} - b_{c'} \equiv \max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} + b_c - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} - b_{c'}
\end{gather}
Ignoring the bias terms, our new optimization objective is:
\begin{gather}
	\max_{\theta} \mathcal{\bm{A}}^{\text{contrastive}} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information as an activation map instead of flattening the feature map. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.

For the multiclass case, our setup now considers a set of classes $c$; $c_t$ is our target class, and $c_{t'}$ consists of a sampled non-target class. The definition for multiclass contrastive activation maps now results in a set of activation maps, defined as follows:
\begin{gather}
	\mathcal{\bm{A}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_t'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
\end{gather}
This is a generalization of the binary case, as the same formula applied to a binary classification challenge returns a set of contrastive maps containing a single element.

To ensure that the model accurately predicts the target class, each contrastive comparison should yield a positive result. Therefore, the multiclass optimization objective under this paradigm is as follows:
\begin{gather}
	\max_{\theta} \tilde{\mathcal{\bm{A}}}_{i}^{\text{contrastive}}\ \forall i \in \mathbb{N}(|c| - 1)
\end{gather}

@@

Segmentations maps describe the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
\begin{gather}
	\{(x, (h, y))\}^N \in \mathcal{D} \text{ where } h_{i,j} := \begin{cases}
		1 & \text{pixel contains subject} \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{background}}_{c,i,j} := \begin{cases}
		\mathcal{\bm{A}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{foreground}}_{c,i,j} := \begin{cases}
		0 & h_{i,j} = 0 \\
		\mathcal{\bm{A}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
	\end{cases} \\
	D_{KL}(P || Q) := \frac{1}{|c|-1}\sum^{|c|-1}_c P \cdot \log \frac{P'}{Q_c} \text{ where } 
	P'_{i,j} := \begin{cases}
		1 & h_{i,j} \neq 0 \\
		P_{i,j} & \text{otherwise}
	\end{cases} \label{d_kl} \\
	\begin{split}
		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 h)) \\
		&+ \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_F \\
		&- \frac{1}{D_1 D_2} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
	\end{split}
\end{gather}

@@

For binary classification, the model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $32$ on a 32GB V100 on a randomly shuffled dataset using the original data splits. The total time for training was $\approx$1hr.

@@

\begin{table}[h]
	\centering
	\begin{tabular}{c|ccc}
		\toprule
		\textbf{Method}  & \textbf{Cross Entropy Loss}  & \textbf{Training Accuracy}   & \textbf{Validation Accuracy} \\
		\midrule
		Cross-Entropy & 0.628 & 68.2\% & 64.1\% \\
		Interpretable (Ours) & \bf 0.327 & \bf 98.4\% & \bf 90.0\% \\
		\bottomrule
	\end{tabular}
\end{table}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=.8\textwidth]{img/plots.png}
	\caption{Plots comparing benchmark cross-entropy loss, training and validation accuracy. The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_dd3fe52afeca42539f87d5da6ac4ea3f\%22,\%22m_e21074d967b741dfb772746b113fe761\%22]}{here}. The orange line represent cross-entropy loss, while the blue line represents the contrastive approach.}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=.4\textwidth]{img/default_cam.png}
	\hspace{4em}
	\includegraphics[width=.4\textwidth]{img/contrastive_cam.png}
	\caption{Left: contrastive activation maps from the default training run. Right: contrastive activation maps from the contrastive training run.}
\end{figure}

@@

For multiclass classification, the objective is to distinguish between the features of species within the cat and dog families. There are a total of 37 classes, with a success of rate of $\approx 2.7\%$ with random guessing.

>>>>

\title{Contrastive Optimization -- Preliminaries}

@@

Where $s_c$ denotes the logit for a single class. $\tilde{\mathcal{\bm{A}}}_c^{\text{HiResCAM}}$ generates a class-specific CAM. For the CAM architecture, \cite{draelos2020use} (section 3.6.2) shows that:
\begin{gather}
	\sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} \\
	s_c = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} + b_c = \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c
\end{gather}
Therefore, we can recreate the entire input-dependent part of the score using HiResCAMs. Doing this for each class $c$, we obtain the bias-removed logit vector $\in \mathbb{R}^{c}$.

@@

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}} - b_{c'}
\end{gather}
Ignoring the bias terms, our new optimization objective is:
\begin{gather}
	\max_{\theta} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}} := \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information as an activation map instead of flattening the feature map. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.

For the multiclass case, our setup now considers a set of classes $c$; $c_t$ is our target class, and $c_{t'}$ consists of a sampled non-target class. The definition for multiclass contrastive activation maps now results in a set of activation maps, defined as follows:
\begin{gather}
	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_t'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
\end{gather}
This is a generalization of the binary case, as the same formula applied to a binary classification challenge returns a set of contrastive maps containing a single element.

To ensure that the model accurately predicts the target class, each contrastive comparison should yield a positive result. Therefore, the multiclass optimization objective under this paradigm is as follows:
\begin{gather}
	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{contrastive}}\ \forall i \in \mathbb{N}(|c| - 1)
\end{gather}

@@

Segmentations maps describe the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
\begin{gather}
	\{(x, (h, y))\}^N \in \mathcal{D} \text{ where } h_{i,j} := \begin{cases}
		1 & \text{pixel contains subject} \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{background}}_{c,i,j} := \begin{cases}
		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{foreground}}_{c,i,j} := \begin{cases}
		0 & h_{i,j} = 0 \\
		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
	\end{cases} \\
	D_{KL}(P || Q) := \frac{1}{|c|-1}\sum^{|c|-1}_c P \cdot \log \frac{P'}{Q_c} \text{ where } 
	P'_{i,j} := \begin{cases}
		1 & h_{i,j} \neq 0 \\
		P_{i,j} & \text{otherwise}
	\end{cases} \label{d_kl} \\
	\begin{split}
		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 h)) \\
		&+ \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_F \\
		&- \frac{1}{D_1 D_2} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
	\end{split}
\end{gather}

@@

For binary classification, the model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $44$ on a 32GB V100 on a randomly shuffled dataset using the original data splits. The total time for training was $\approx$1hr.

@@

\begin{table}[h]
	\centering
	\begin{tabular}{c|ccc}
		\toprule
		\textbf{Method}  & \textbf{Cross Entropy Loss}  & \textbf{Training Accuracy}   & \textbf{Validation Accuracy} \\
		\midrule
		Cross-Entropy & 0.646 & 66.1\% & 64.5\% \\
		Interpretable (Ours) & \bf 0.326 & \bf 98.6\% & \bf 92.3\% \\
		\bottomrule
	\end{tabular}
\end{table}

@@

\vspace{-2em}
\begin{figure}[H]
	\centering
	\includegraphics[width=.4\textwidth]{img/accuracy.png}
	\hspace{4em}
	\includegraphics[width=.4\textwidth]{img/loss.png}
	\caption{Plots comparing benchmark cross-entropy loss, training and validation accuracy. The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_f4f7dea52fa74739be788e5f47da1195\%22,\%22m_1301bf1ba9bf4db48e09e7497c8cd214\%22]}{here}. The blue line represents cross-entropy loss, while the orange line represents the contrastive approach.}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=.4\textwidth]{img/default_cam.png}
	\hspace{3em}
	\includegraphics[width=.4\textwidth]{img/contrastive_cam.png}
	\caption{Left: contrastive activation maps from the default training run. Right: contrastive activation maps from the contrastive training run.}
\end{figure}

@@

For multiclass classification, the objective is to distinguish between the features of species within the cat and dog families. There are a total of 37 classes, with a success of rate of $\approx 2.7\%$ with random guessing. There is virtually no class imbalance in this case.
